---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
library(stringr)
```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
mdexpenses <- read_csv("data/maryland_expenses.csv")

mdexpenses
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

   1. Create a dataframe of Media expenses where the name of the payee is not NA.
   2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.** Friends of Jain Ashwani paid the most to Facebook at $62,357.38	 

```{r}
cleanmd <- mediaexpenses |> 
  clean_names()

mediaexpenses <- cleanmd |> 
  filter(expense_category == "Media")

cleanmedia <- mediaexpenses |> 
  filter(!is.na(payee_name))

write_csv(cleanmedia,"data/cleanmedia.csv")
```

```{r}
refinedmedia <- read_csv("data/refinedmedia.csv")
```


Whoops forgot to clean the dates earlier
```{r}
finalformmedia <- refinedmedia |> 
 mutate(expenditure_date=mdy(expenditure_date))


view(finalformmedia)
```

Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

```{r}
finalformmedia |> 
  filter(payee_clean == "Facebook") |> 
group_by(committee_name) |>
  summarise(totalamount = sum(amount)) |> 
arrange (desc(totalamount)) 
view(finalformmedia)
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** Of the top 10 spenders, eight are candidate committies for MD governor, county executive or county council. The other two are Casa in Action PAC and WFP National PAC. The former is a Latino and immigrant organization that endorses candidates that align with its values and promotes voting and the latter deals with issues concerning healthcare reform, raising minimum wage, public education, etc. The ad information over Facebook shows how much was spent on the ads, when they ran, the estimated audience size, category, impressions, platforms, give it a Library ID and list if they're active or inactive.


**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** Wes Moore and his running mate Aruna Miller are missing from the Q1 list, as well as the Republican primary winner Dan Cox and his running mate Gordana Schifanelli. Both campaigns ran Facebook ads (according to the ad search page. Wes Moore doesn't have Facebook listed here). The biggest recepients of media expenses by Wes Moore were to SKDK, which is a public affairs and political consulting firm, Foglamp, which looks to specialize in digital strategies, and Apollo Artistry, which designs platforms, logos and content strategies for clients.

```{r}
#Find how much Wes Moore and his joint committee spent on media expenses
#I had to use ChatGPT just to remember the | is better than the comma in this situation. Search was "how to filter for two values in the same column in r tidyverse explain like I'm new to code"

finalformmedia |> 
  filter(committee_name == "Moore  Wes For Maryland" | committee_name == "Miller  Aruna Friends of" | committee_name == "Moore Miller For Maryland Slate")|>
group_by(payee_clean) |> 
summarise(totalamount = sum(amount)) |> 
arrange (desc(totalamount))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** It's vastly different, although Casa in Action PAC does show up in the top 5 here. Moore Miller for Maryland also shows up in the top 10. Of the top 5 committees who spent the most on Facebook ads, Franchot Anderson-Walker for Maryland Slate and Friends of Ashwani Jain appeared to have spent all of their online advertising dollars on Facebook. In the top 10, Friends of Jasani Salimah joins that list.

```{r}
onlineads <- finalformmedia |> 
  filter(expense_purpose == "Online Advertising") |> 
  group_by(committee_name) |>
  summarise(totalamount = sum(amount)) |> 
  arrange (desc(totalamount)) 
view(onlineads)
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** We know Wes Moore spent most of his media budget on SKDK and a select few other firms directly. We an also see that the campaigns that only spent their media money on Facebook ads didn't do very well. 

When I looked for Wes Moore over the Facebook ad library, it showed that his campaign had paid for an ad. However, it didn't show up in this data. It's possible that he created it after this database was created, but I'd be interested in following that. Another story might be to see how many politicians running for office for the first time spend all of their online media advertisement money on Facebook ads. I would also love to know how much SKDK makes from Democrats running for office and if the top spenders are elected more often than not.
