---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
library(dplyr)
library(readr)
library(stringr)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
```
## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
ACS5 <- load_variables(2021, "acs5", cache = TRUE)
view(ACS5)

```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 
**A1** Prince George's County

```{r}
md <- get_acs(geography = "county",
               variables = c(medianage = "B01002_001"),
               state = "MD",
               year = 2021)

md_arrange <- md |> 
  arrange((estimate)) |>  
  top_n(20, estimate)

```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 
**A2** Talbot County; Worcester County

```{r}
at_least_50 <- md |> 
  filter(estimate >= 50)

at_least_50

```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1. Load the Montgomery County 911 overdose calls data.
```{r}
montgomery2022 <- read_csv("data/montgomery_2022_overdoses.csv") |> 
mutate(zip = as.character(zip))
view(montgomery2022)
```
2. Create a new dataframe with the total number of calls for each zip code.

```{r}
mont2022 <- montgomery2022 |> 
group_by(zip) |> 
summarise(count = n()) |> 
arrange(desc(count)) |> 
mutate(zip = as.character(zip))

mont2022

```
3. Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one). 

#B16003_008 -- this was one of the only two provided, but the concept was odd. I asked ChatGPT and it gave me a variable that wasn't in this dataset. I don't understand why we need the total 18 and up population instead of the whole population. There are a lot of parts to this question and it got really confusing as to why we were doing each part/what to use -- I feel like I have too many datasets. Here's my convo with Chat GPT https://chat.openai.com/c/2410502c-3a23-4dff-95b1-fd181835e213

4. Get from the ACS data from all Maryland zip codes using that variable.

```{r}
mdzip <- get_acs(geography="zcta", variables = "B16003_008", state="MD", year=2019)
mdzip
```

5. Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.

```{r}

mont_merge <- montgomery2022 |> 
  left_join(mdzip, by=c('zip'='GEOID')) |> 
  rename(population = estimate)

mont_merge
```

6. Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

```{r}
mont_per_capita <- mont_merge |> 
  group_by(zip, population, moe) |> 
  summarise(count=n()) |> 
mutate(per_capita = (count/population)*1000) |> 
  arrange(desc(per_capita))
  
mont_per_capita
```
Which zip code has the highest rate of calls? Which has the lowest?

**A3** 20872 had the highest rate of calls. 20707 had the lowest.

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4** I'm a bit wary to use this since I don't think I grabbed the right variable for 18+. For the demographics, over half of the population in the zip code 20872 is white, with 20% of the population Hispanic. It also has a generally older median age than the rest of Maryland at 54.9 but it also has a pretty low population in general. In comparison, the median age in the zip code 20707 is 38. The zip code has a large Black population. 8.4% of the people here are below the poverty line compared to .8% of persons in the 20872 zip code. That could be a story, but also I would want to double check it with the data from the correct variable.

**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5** I don't think I did this right. For the max population, 20861 took first with 20872, my original zip code, in second. The first zip code only has 12 people in it supposedly though. 20707 still has the lowest per_capita rate. I'm confident that I might have the done the code correctly but not confident in the answers as I probably used the wrong variable.


```{r}
max_pop <- mont_merge |> 
  mutate(min = population - moe, max = population + moe) |> 
  group_by(zip, max) |> 
  summarise(count=n()) |> 
  mutate(per_capita = (count/max)*1000) |> 
  arrange(per_capita)
max_pop
```
```{r}
min_pop <- mont_merge |> 
  mutate(min = population - moe, max = population + moe) |> 
  group_by(zip, max) |> 
  summarise(count=n()) |> 
  mutate(per_capita = (count/max)*1000) |> 
  arrange(desc(per_capita))
min_pop
```



```{r}
#I had to consult Chat GPT for this -- there were a few typos to fix
graph <- mont_merge |> 
  mutate(min = population - moe, max = population + moe) |> 
  ggplot(aes(x = population, y = zip)) +
  geom_errorbar(aes(xmin = population - moe, xmax = population + moe)) +
  geom_point(color = "red") +
  labs(title = "Montgomery County overdoses by zip code",
       subtitle = "2017-2021 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")

graph

```

