---
title: "lab_09.Rmd"
author: "derek willis"
date: "2023-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Datawrapper account
* Data in CSV files in the `data` folder

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.
```{r}
install.packages("scales")
```

```{r}
# load libraries here
library(tidyverse)
library(janitor)
library(dplyr)
library(scales)
```

## Questions

**Q1.** You want to explore the differences between the electoral performances of Dan Cox in 2022 and Larry Hogan in 2018. Except a map is probably not the best way to show that. Let's make a chart instead, and the type is up to you, but the goal is to clearly show the differences between the two.

You MUST do the following things when answering this question:
* Load the md_gov_county.csv and md_gov_county_18.csv files from the data folder
* Join those dataframes, specifying the join column. 
* Add columns to generate percentages for Cox and Hogan.
* Export that joined dataframe to a CSV file using `write_csv`.
* Using that exported CSV, make a chart with Datawrapper. What kind of chart is up to you, so experiment with a few kinds of charts to see what answers the question above in the clearest way. Your chart MUST:
  * Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.
  * Easily readable labels and values
* Publish your chart
* Describe your thought process below - why you chose to make the chart you did - and add the URL of your chart.

**A1.** https://datawrapper.dwcdn.net/Mqkbe/3/
At first I wanted to use vertical columns but the data was giving me trouble with that. It kept grouping them by percent of votes instead of by county and I couldn't transpose the data. In the end, I think this visualization still works and is still pretty impactful. I thought about trying to use a map instead, but I like how this shows the gap between voters in each election, which is something a map wouldn't be able to do. This also helped me see that there was data missing from Queen Anne's county, so I went back into the code and added it. For the title, I didn't want to outright say that voter turnout was why Cox's numbers were lower because I couldn't account for what part was from voter turn out v. Republicans voting for a third party or even Democrat, but I did some research and there was poor voter turnout that year. Also, looking at the total votes for the Republican candidate in 2022 and 18, there is a vast gap between them compared to the total votes for the Democrat candidate in both years. I think it also would have been interesting to calculate the percent change for each year by county as a separate chart.

```{r}
#Here I am cleaning the datasets and renaming the "write_ins" column of each data set so I don't confuse the two or confuse R
clean_md_gov_county <- read_csv("data/md_gov_county.csv") |> 
  clean_names() |> 
rename(write_ins_2022 = write_ins)

clean_md_gov_county_18 <- read_csv("data/md_gov_county_18.csv") |> 
  clean_names() |> 
rename(write_ins_2018 = write_ins)

#Here I am joining the two datasets using left_join() by the geoid and county

joined_years <- left_join(clean_md_gov_county, clean_md_gov_county_18, by = c("geoid", "county"))

#Uh-oh, Saint Mary's County doesn't have data for 2018
#Source: https://elections.maryland.gov/elections/2018/results/general/gen_results_2018_2_by_county_19-1.html

all_counties <- joined_years |> 
  mutate(hogan = ifelse(is.na(hogan), 30703, hogan),
         jealous = ifelse(is.na(jealous), 9084, jealous),
         quinn = ifelse(is.na(quinn), 410, quinn),
         schlakman = ifelse(is.na(schlakman), 170, schlakman),
         write_ins_2018 = ifelse(is.na(write_ins_2018), 27, write_ins_2018)
         )

all_counties
```

```{r}
#Here I'm adding the total and percentage columns while also taking out the other candidates to get rid of any noise.

hogan_v_cox <- all_counties |> 
  mutate(total_2022 = cox + moore + lashar + wallace + harding + sefcik + write_ins_2022) |> 
  mutate(pct_cox = cox/total_2022 * 100) |> 
  mutate(total_2018 = hogan + jealous + quinn + schlakman + write_ins_2018) |> 
  mutate(pct_hogan = hogan/total_2018 * 100) |> 
  select(county, cox, hogan, pct_hogan, pct_cox)

write_csv(hogan_v_cox, "data/hogan_v_cox.csv")
```


```{r}
#I wanted to see if I could tell if it was due to voter turnout, but I don't think I can confirm with just data/no survey. The state board of election office did confirm that the 2022 elections had poor turnout in general, however.

total_hogan_v_cox <- all_counties |> 
  mutate(total_2022 = cox + moore + lashar + wallace + harding + sefcik + write_ins_2022) |> 
  mutate(pct_cox = cox/total_2022 * 100) |> 
  mutate(total_2018 = hogan + jealous + quinn + schlakman + write_ins_2018) |> 
  mutate(pct_hogan = hogan/total_2018 * 100) |> 
  select(geoid, county, cox, hogan, jealous, moore, total_2018, total_2022, pct_hogan, pct_cox) |> 
    summarise(sum(total_2018),
              sum(total_2022),
              sum(hogan),
              sum(cox),
              sum(jealous),
              sum(moore))

total_hogan_v_cox
```

**Q2.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A2.** https://datawrapper.dwcdn.net/VglwT/1/
Voting on election day bounced back in the 2022 election after the percentage of that voting type plummeted in 2020 -- during the pandemic when mail-in voting was encouraged. While levels aren't back to how they were before COVID, it's by far the most popular way to vote once again. Another story that stood out to me was the total of votes cast when comparing the years. I know 2020 was a presidential election, but the 2022 total votes were also lower than the 2018 votes. Again, poor turnout as seen in the last question.

```{r}

vote_types <- read_csv("data/md_votes_by_type.csv") |> 
  clean_names() |> 
  mutate(early_percent = (early/total) * 100,
         election_percent = (election_day/total) * 100,
         mail_percent = (mail/total) * 100) 

write_csv(vote_types, "data/vote_types.csv")

```

**Q3.** You want to know more about the geography of 911 overdose calls in Montgomery County, Maryland, in 2022. To do that, you'll make a choropleth map in Datawrapper. You will need to:

* Read the file into R and create columns for the number of calls in each zip code and the percentage of all calls that number represents and then export that CSV to use in your map.
* Search for Maryland zip codes (use the 2020 data) as the base map.
* Upload the CSV file.
* Change the tooltip to show both the number and the percentage (you can put them side by side or put one under the other). Make sure to properly format both numbers to make them readable.
* Add labels to your map that locate prominent places in the county.
* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.
* Make any other changes to data that helps you make a more accurate map.

What do you think the story here is, and what would you want to know more about? Put the answer below along with the URL of your map.

**A3.** https://datawrapper.dwcdn.net/mEo0P/1/
I think there might be a story with Germantown and why there are so many overdoses in that zip code and a few neighboring ones. Is it population? It doesn't feel like that's the whole picture because Silver Spring's numbers aren't as bad as Georgetown. It might lie in less access to healthcare since that seemed to be a big issue on its Census page. Are people not calling the ambulance because the person who overdosed can't afford care? There's also a large Hispanic population, so it's possible there are some people who are undocumented and they don't want to call for care. I would be hesitant to touch this story without careful research and guidance though if that is a story as I wouldn't want to endanger someone/risk someone being taken from their family. It would be a good story on access to care for a vulnerable population though if done carefully and respectfully.

#I had to ask ChatGPT for help on formatting the percent column since that was annoying me in datawrapper. It ended up not transfering the percent sign, so I had to look up the code in html (&percnt;)
```{r}
mont_overdoses <- read_csv("data/montgomery_2022_overdoses.csv") |> 
  clean_names() |> 
#  separate(start_time, into = c("date", "time"), sep = " ") |> 
#  arrange(desc(date))
#This is from when I wanted to double check everything was from 2022
  group_by(zip) |> 
  summarise(calls_per_zip = n()) |> 
  arrange(desc(calls_per_zip)) |> 
  mutate(total_calls = sum(calls_per_zip)) |> 
  mutate(percent_of_calls = (calls_per_zip / total_calls) * 100) |> 
  mutate(percent_of_calls_formatted = scales::percent(percent_of_calls, scale = 1, suffix = "%", accuracy = 0.01)) |>
  select(zip, calls_per_zip, total_calls, percent_of_calls_formatted)

write_csv(mont_overdoses, "data/montgomery_overdoses.csv")
```
